\bf Distortion corrected kernel density estimate on Riemannian manifolds

### 2023-06-01 jcgs/kderm_jcgs.tex
Add jcgs version for kderm, the latest version

\documentclass[11pt,a4paper,]{article}
\usepackage{lmodern}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Distortion corrected kernel density estimate on Riemannian manifolds},
            pdfkeywords={manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\title{Distortion corrected kernel density estimate on Riemannian manifolds}

%% MONASH STUFF

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

%% FONT
\RequirePackage{bera}
\RequirePackage{mathpazo}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
\makeatletter
\lhead{\textsf{\expandafter{\@title}}}
\makeatother
\rhead{}
\cfoot{}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% MATHS
\RequirePackage{bm,amsmath}
\allowdisplaybreaks

%% GRAPHICS
\RequirePackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.8}

%\RequirePackage[section]{placeins}

%% SECTION TITLES
\RequirePackage[compact,sf,bf]{titlesec}
\titleformat{\section}[block]
  {\fontsize{15}{17}\bfseries\sffamily}
  {\thesection}
  {0.4em}{}
\titleformat{\subsection}[block]
  {\fontsize{12}{14}\bfseries\sffamily}
  {\thesubsection}
  {0.4em}{}
\titlespacing{\section}{0pt}{*5}{*1}
\titlespacing{\subsection}{0pt}{*2}{*0.2}

%% TITLE PAGE
\def\Date{\number\day}
\def\Month{\ifcase\month\or
 January\or February\or March\or April\or May\or June\or
 July\or August\or September\or October\or November\or December\fi}
\def\Year{\number\year}

\makeatletter
\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
\def\nojel{\def\showjel{}}
\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
\def\cover{{\sffamily\setcounter{page}{0}
        \thispagestyle{empty}
        \placefig{2}{1.5}{width=5cm}{monash2}
        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
        \begin{textblock}{7}(12.7,27.9)\hfill
        \includegraphics[height=0.7cm]{AACSB}~~~
        \includegraphics[height=0.7cm]{EQUIS}~~~
        \includegraphics[height=0.7cm]{AMBA}
        \end{textblock}
        \vspace*{2cm}
        \begin{center}\Large
        Department of Econometrics and Business Statistics\\[.5cm]
        \footnotesize http://monash.edu/business/ebs/research/publications
        \end{center}\vspace{2cm}
        \begin{center}
        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
                \LARGE\@author\end{onehalfspace}
        }}
        \end{center}
        \vfill
                \begin{center}\Large
                \Month~\Year\\[1cm]
                Working Paper \@wp
        \end{center}\vspace*{2cm}}}
\def\pageone{{\sffamily\setstretch{1}%
        \thispagestyle{empty}%
        \vbox to \textheight{%
        \raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
        \vspace{2cm}\par
        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
        \hspace{1cm}\showjel\vss}}}
\def\blindtitle{{\sffamily
     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
        }}
\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}

\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
\let\maketitle\titlepage
\makeatother

%% SPACING
\RequirePackage{setspace}
\spacing{1.5}

%% LINE AND PAGE BREAKING
\sloppy
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% PARAGRAPH BREAKS
\setlength{\parskip}{1.4ex}
\setlength{\parindent}{0em}

%% HYPERLINKS
\RequirePackage{xcolor} % Needed for links
\definecolor{darkblue}{rgb}{0,0,.6}
\RequirePackage{url}

\makeatletter
\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
\makeatother
\hypersetup{
     citecolor=0 0 0,
     breaklinks=true,
     bookmarksopen=true,
     bookmarksnumbered=true,
     linkcolor=darkblue,
     urlcolor=blue,
     citecolor=darkblue,
     colorlinks=true}

%% KEYWORDS
\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}

%% ABSTRACT
\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
  {\end{minipage}}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage[showonlyrefs]{mathtools}
\usepackage[no-weekday]{eukdate}

%% BIBLIOGRAPHY

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}

\DeclareFieldFormat{url}{\texttt{\url{#1}}}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
\DeclareFieldFormat[inproceedings]{title}{#1}
\DeclareFieldFormat{shorthandwidth}{#1}
% No dot before number of articles
\usepackage{xpatch}
\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
% Remove In: for an article.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}

\makeatletter
\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
\makeatother
\renewcommand*{\finalnamedelim}{%
  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
  \addspace\&\space}

\nojel

\RequirePackage[absolute,overlay]{textpos}
\setlength{\TPHorizModule}{1cm}
\setlength{\TPVertModule}{1cm}
\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}

\nocover

\author{Fan~Cheng, Anastasios~Panagiotelis, Rob J~Hyndman}
\addresses{\textbf{Fan Cheng}\newline
Monash University
\newline{Email: \href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}}\\[1cm]
\textbf{Anastasios Panagiotelis}\newline
The University of Sydney
\newline{Email: \href{mailto:Anastasios.Panagiotelis@sydney.edu.au}{\nolinkurl{Anastasios.Panagiotelis@sydney.edu.au}}}\\[1cm]
\textbf{Rob J Hyndman}\newline
Monash University
\newline{Email: \href{mailto:Rob.Hyndman@monash.edu}{\nolinkurl{Rob.Hyndman@monash.edu}}}\\[1cm]
}

\date{\sf\Date~\Month~\Year}
\makeatletter
 \lfoot{\sf Cheng, Panagiotelis, Hyndman: \@date}
\makeatother

%% Any special functions or other packages can be loaded here.
\usepackage{tabu,placeins,algorithmic}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\urlstyle{tt}  % use monospace font for urls
\usepackage[cal = txupr]{mathalpha}
\geometry{margin=2.2cm}
\mathtoolsset{showonlyrefs}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
% \def\equationautorefname~#1\null{Equation~(#1)\null} % NOT showing Equation 1, but Section 2.1
\usepackage{bm}
\DeclareMathOperator\supp{supp}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

% Adjust headwidth in case user has changed geometry in header-includes
\renewcommand{\headwidth}{\textwidth}

\begin{document}
\maketitle
\begin{abstract}
Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with a 2-D data from Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space. We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data.
\end{abstract}
\begin{keywords}
manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
\end{keywords}

\newpage

@@

Multivariate kernel density estimation (KDE see \textcite{Parzen1962-gt}; \textcite{Chen2017-dw} and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment \autocite{Xie2008-eb,Okabe2009-nb}, automatic detection in visual surveillance systems \autocite{Elgammal2002-cw}, wind power density detection \autocite{Jeon2012-ac} and prime prediction via Twitter messages \autocite{Gerber2014-tq}. For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\). The problem of KDE on manifolds has been well studied in \textcite{Hendriks1990-bt}, \textcite{Izenman1991-cj}, \textcite{Pelletier2005-vu}, \textcite{Henry2013-da} and \textcite{Berenfeld2022-fd}. However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. Our objective in this paper is to propose KDE for data that lie on some unknown manifold.

When dealing with data that lie on a manifold, it is common to use dimension reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\). Details of these algorithms are summarized in \textcite{Cayton2005-dp}, \textcite{Lee2007-wq}, \textcite{Izenman2012-mx} and \textcite{Cheng2021-dh}. One could easily apply KDE to the output vectors \(\pmb{y}\). However, this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,\]
where \(f\) is the probability density function on the manifold, \(\pmb{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\). By a change of variables, this would imply the following density for \(\pmb{y}\),
\begin{equation}
\label{eq:changevar}
Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
\end{equation}
where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\pmb{H}(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\). It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|\pmb{H}(\pmb{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant. Since \(\psi\) will typically be a non-linear function, the value of \(|\pmb{H}(\pmb{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\pmb{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm. For this reason, we refer to it as the distortion-corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq} which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is of anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data. Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic \autocite[see Section 6.6 of][ for a summary]{Scott2015-vl}.

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. Our estimator combines insights from two existing papers, the first is \textcite{Pelletier2005-vu} who propose a KDE for data on manifolds that are known, the second is \textcite{Perrault-Joncas2013-pq} who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm. \autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space. \autoref{dckdeapplication} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage. Conclusions and discussions are presented in \autoref{dckdeconclusion}. Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.

@@

In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of \textcite{Pelletier2005-vu}, which can be applied when the data lie on some known manifold. In \autoref{MetLearn}, we describe the Learn Metric algorithm of \textcite{Perrault-Joncas2013-pq}, which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of \textcite{Pelletier2005-vu} and \textcite{Perrault-Joncas2013-pq}, we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

%% Any special functions or other packages can be loaded here.
\usepackage{tabu,placeins,algorithmic}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\urlstyle{tt}  % use monospace font for urls
\usepackage[cal = txupr]{mathalpha}
\usepackage{geometry}
\geometry{margin=2.2cm}
\usepackage{mathtools}
\mathtoolsset{showonlyrefs}
\usepackage{amsfonts,amssymb}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
% \def\equationautorefname~#1\null{Equation~(#1)\null} % NOT showing Equation 1, but Section 2.1
\usepackage{bm}
\DeclareMathOperator\supp{supp}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Distortion corrected kernel density estimate on Riemannian manifolds}

  \author{
        Fan Cheng \thanks{\href{mailto:Fan.Cheng@monash.edu}{\nolinkurl{Fan.Cheng@monash.edu}}} \\
    Monash University\\
     and \\     Rob J Hyndman \\
    Monash University\\
     and \\     Anastasios Panagiotelis \\
    The University of Sydney\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Distortion corrected kernel density estimate on Riemannian manifolds}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Manifold learning can be used to obtain a low-dimensional representation of the underlying Riemannian manifold given the high-dimensional data. However, kernel density estimates of the low-dimensional embedding with a fixed bandwidth fail to account for the way manifold learning algorithms distort the geometry of the Riemannian manifold. We propose a novel distortion-corrected kernel density estimator (DC-KDE) for any manifold learning embedding by introducing the estimated Riemannian metric of each point to fix the distortion in the line and volume elements. The geometric information of the manifold guarantees a more accurate density estimation of the true manifold, which subsequently could be used for anomaly detection. To compare our proposed estimator with a fixed-bandwidth kernel density estimator, we run two simulations with 2-D data from the Gaussian mixture model mapped into a 3-D twin peaks shape and a 5-D semi-hypersphere mapped in a 100-D space. We demonstrate that the proposed DC-KDE could improve the density estimates given a good manifold learning embedding and has higher rank correlations with the true manifold density. A shiny app in R is also developed for various simulation scenarios. The proposed method is applied to density estimation in statistical manifolds of electricity usage with the Irish smart meter data. This demonstrates the estimator's capability to fix the distortion of the manifold geometry and a new approach to anomaly detection in high-dimensional data.
\end{abstract}

\noindent%
{\it Keywords:} manifold learning, variable bandwidth, Riemannian metric, geodesic distance, Gaussian kernels
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

@@

Multivariate kernel density estimation (KDE see Parzen (\protect\hyperlink{ref-Parzen1962-gt}{1962}); Scott (\protect\hyperlink{ref-Scott2015-vl}{2015}) and references therein) is a fundamental tool for exploratory data analysis and unsupervised learning. Applications of KDE include finding hot spots of traffic networks in the GIS environment (\protect\hyperlink{ref-Xie2008-eb}{Xie and Yan 2008}; \protect\hyperlink{ref-Okabe2009-nb}{Okabe, Satoh, and Sugihara 2009}), automatic detection in visual surveillance systems (\protect\hyperlink{ref-Elgammal2002-cw}{Elgammal et al. 2002}), wind power density detection (\protect\hyperlink{ref-Jeon2012-ac}{Jeon and Taylor 2012}) and prime prediction via Twitter messages (\protect\hyperlink{ref-Gerber2014-tq}{Gerber 2014}). For \(m\)-dimensional data where \(m\) is even of moderate size, KDE is challenging due to the well-known curse of dimensionality. However, the manifold hypothesis suggests that many high-dimensional datasets belong to a \(d\)-dimensional manifold \(M\) where \(d\ll m\). The problem of KDE on manifolds has been well studied in Hendriks (\protect\hyperlink{ref-Hendriks1990-bt}{1990}), Izenman (\protect\hyperlink{ref-Izenman1991-cj}{1991}), Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}), Henry, Muñoz, and Rodríguez (\protect\hyperlink{ref-Henry2013-da}{2013}), and Berenfeld, Rosa, and Rousseau (\protect\hyperlink{ref-Berenfeld2022-fd}{2022}). However, in these settings, it is assumed that the underlying manifold is known, which is rarely true in practice. Our objective in this paper is to propose KDE for data that lie on some unknown manifold.

When dealing with data that lie on a manifold, it is common to use dimension-reduction techniques. These algorithms which include ISOMAP, LLE, t-SNE, and UMAP among others, can be thought of as a mapping \(\psi:M\rightarrow \mathbb{R}^d\) that take points on the manifold \(\pmb{p}\) to \(d\)-dimensional vectors \(\pmb{y}\). Details of these algorithms are summarized in Cayton (\protect\hyperlink{ref-Cayton2005-dp}{2005}), Lee and Verleysen (\protect\hyperlink{ref-Lee2007-wq}{2007}), Izenman (\protect\hyperlink{ref-Izenman2012-mx}{2012}), and Cheng, Panagiotelis, and Hyndman (\protect\hyperlink{ref-Cheng2021-dh}{2021}). One could easily apply KDE to the output vectors \(\pmb{y}\). However, this would be a misleading representation of the density of points as they lie on the manifold. To see how, consider that the density on the manifold must be defined with respect to an appropriate volume form, that is
\[Pr(\pmb{p}\in\mathcal{A})=\int_\mathcal{A} f(\pmb{p}) dvol_M,\]
where \(f\) is the probability density function on the manifold, \(\pmb{p}\) is a point on the manifold, \(\mathcal{A}\) is some region on the manifold and \(dvol_M\) is a volume form on the manifold \(M\). By a change of variables, this would imply the following density for \(\pmb{y}\),
\begin{equation}
\label{eq:changevar}
Pr(\pmb{y}\in\psi(\mathcal{A}))=\int\limits_{\psi(\mathcal{A})} (f\circ\psi^{-1})(\pmb{y})|\pmb{H}(\pmb{y})|^{1/2}d\pmb{y},
\end{equation}
where \(\psi(\mathcal{A})\) is the image of \(\mathcal{A}\) under \(\psi\) and \(\pmb{H}(\pmb{y})\) is the Riemannian metric expressed in local coordinates given by the mapping \(\psi\). It can be seen that the density in terms of \(\pmb{y}\), differs from the density on the manifold due to the term \(|\pmb{H}(\pmb{y})|^{1/2}\) which can be considered roughly analogous to the Jacobian determinant. Since \(\psi\) will typically be a non-linear function, the value of \(|\pmb{H}(\pmb{y})|^{1/2}\) will change across the manifold, in a way that measures the distortion induced by the dimension reduction algorithm. Notably, different choices of dimension reduction algorithm, imply different \(\psi\), meaning that if KDE is applied to the output \(\pmb{y}\) coming from two different dimension reduction algorithms, this could lead to two very different sets of conclusions about the density of the same point.

In this paper, we propose a new KDE. Our estimator is also based on the output of dimension reduction algorithms but accounts for the distortion induced by the algorithm. For this reason, we refer to it as the distortion-corrected KDE (DC-KDE). Critical to our estimator, is obtaining an estimate of \(|\pmb{H}(\pmb{y})|^{1/2}\) for this purpose we use the Learn Metric algorithm of Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}) which augments any dimension reduction algorithm with an estimate of the Riemannian at each data point. One particular downstream task that we are interested in is anomaly detection, where anomalies are identified as points of low density. In a number of simulation studies, we demonstrate two benefits that arise when using our new estimator. The first is that using distortion correction leads to more accurate density estimates and better detection of anomalies compared to the case where distortion correction is not used. The second is that when the estimates are found using different dimension reduction algorithms (e.g.~ISOMAP v UMAP), the results regarding the density estimates and the detection of anomalies are robust only when distortion correction is applied.

In terms of data visualization, our approach allows a two or three-dimensional scatterplot of \(\pmb{y}\) to be augmented by coloring points according to the magnitude of the density estimate. In this way, the user is able to tell whether a region that has a cluster of points in the output space \(\pmb{y}\) is truly a high-density region or if instead, this is an artifact of the dimension reduction algorithm `squeezing' points together. This is particularly pertinent in the case of t-SNE which can form clusters of points even when no such clusters exist in the underlying data. Furthermore, our estimator can be used for manifolds with elements that are a variety of objects, for instance, rotational frames or images. In our application we consider a probabilistic manifold whose elements are densities, meaning that we estimate a density of densities. Finally, our proposed estimator also takes the form of a KDE with variable bandwidth meaning that it also contributes to extensive literature on this topic (see Section 6.6 of \protect\hyperlink{ref-Scott2015-vl}{Scott 2015} for a summary).

The rest of the paper is organized as follows. In \autoref{kderm}, we present our distortion-corrected kernel density estimator for Riemannian manifolds. Our estimator combines insights from two existing papers, the first is Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}) who propose a KDE for data on manifolds that are known, the second is Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}) who propose the Learn Metric algorithm which we use to estimate the distortion induced by a dimension reduction algorithm. \autoref{simulation} is composed of two simulations; the first deals with 2-dimensional data from a Gaussian mixture model mapped onto a 3-dimensional `twin peaks' manifold and the second with a 5-dimensional semi-hypersphere data embedded into a 100-dimensional space. \autoref{dckdeapplication} contains an application to an Irish smart meter dataset that visualizes and identifies households with an anomalous distribution of energy usage. Conclusions and discussions are presented in \autoref{dckdeconclusion}. Some additional background on important terms in Riemannian geometry required to understand our proposed method is provided in \autoref{riemgeo}.

@@

In this section, we introduce our method for kernel density estimation on manifolds that uses an embedding from a dimension reduction algorithm while correcting for the distortion induced by this embedding. Since some readers may be unfamiliar with the nuances of manifolds, we first discuss kernel density estimation for data in Euclidean space, then illustrate in \autoref{Pellet} how this generalizes to the estimator of Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}), which can be applied when the data lie on some known manifold. In \autoref{MetLearn}, we describe the Learn Metric algorithm of Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}), which augments an embedding derived from a dimension reduction algorithm with an estimate of the Riemannian metric expressed in local coordinates. By combining elements from the work of Pelletier (\protect\hyperlink{ref-Pelletier2005-vu}{2005}) and Perrault-Joncas and Meila (\protect\hyperlink{ref-Perrault-Joncas2013-pq}{2013}), we derive our own novel distortion-corrected kernel density estimate in \autoref{DCKDE}. To keep this section as succinct as possible, we do not define concepts such as manifolds, charts, geodesic distance, etc., but provide this information for readers unfamiliar with differential geometry in \autoref{riemgeo}.
